---
title: "Using spectator - worked examples"
output: rmarkdown::html_vignette
# output: pdf_document
vignette: >
  %\VignetteIndexEntry{Using spectator - worked examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spectator)
```
## spectator R package examples

In absence of a complete user manual, this document shows some examples of using the "agri4cast3" package. Good luck! 

```{r}
library(agri4cast3)
options(max.print = 20)
```

The function call: GetGridData(rows, cols, period, vars)
```{r}
# GetGridData - function call
# 
a1 <- GetGridData(rows = 81:82, cols = 91:94, period = 19910101:19910630, vars = c("TMIN", "TMAX"))
a1b <- GetGridData(c(81, 82), c(91, 94), c(19910101, 19910630), c("TMIN", "TMAX"))
a1c <- GetGridData(c(81, 82), c(91, 94), c("1991-01-01", "1991-06-30"), c("TMIN", "TMAX"))
jour <- as.Date("1991-01-01")
a1d <- GetGridData(c(81, 82), c(91, 94), c(jour, jour + 180), c("TMIN", "TMAX"))
```

```{r}
print(identical(a1, a1b))
print(identical(a1, a1c))
print(identical(a1, a1d))
```

```{r}
# GetGridData - result
# 
a1 <- GetGridData(81:82, 91:94, 19910101:19910630, c("TMIN", "TMAX"))
```

```{r}
print(length(a1))
print(names(a1))
print(dim(a1[[1]]))
print(dim(a1$TMIN))
print(dimnames(a1[[1]]))
print(a1[[1]])
```

Getting data by years:
```{r}
# GetGridData - get by year
# 
a2 <- GetGridData(81:82, 91:94, 1991:1996, c("TMIN", "TMAX"))
```

```{r}
print(length(a2))
print(names(a2))
print(dim(a2[[1]]))
print(dimnames(a2[[1]]))
```


Getting data by country codes (two-letter ISO 3166-1 alpha-2): GetCountryData(country, period, vars)
```{r}
# GetCountryData
# 
a3 <- GetCountryData("VA", c(19910101, 19910630), c("TMIN", "TMAX"))
```

```{r}
print(length(a3))
print(names(a3))
print(dim(a3[[1]]))
print(dimnames(a3[[1]]))
```


Alternative to keep all dimensions:
```{r}
a3b <- GetCountryData("VA", c(19910101, 19910630), c("TMIN", "TMAX"), drop = FALSE)
```

```{r}
print(length(a3b))
print(names(a3b))
print(dim(a3b[[1]]))
print(dimnames(a3b[[1]]))
```


Getting data by geographic coordinates (latitude, longitude - like in Google Maps): GetPointData(latitude, longitude, period, vars)
```{r}
# GetPointData
# 
a4 <- GetPointData(45.847113, 3.167136, c(19910101, 19910630), c("TMIN", "TMAX"))
```

```{r}
print(length(a4))
print(names(a4))
print(dim(a4[[1]]))
print(dimnames(a4[[1]]))
```


## Crop models

Getting data in format ready to be used with crop models: GetGridCropSim(rows, cols, period)
```{r}
# GetGridCropSim
# 
b1 <- GetGridCropSim(81:82, 91:94, 1991:1996)
```

This is just a shortcut for:
```{r}
b1b <- ReshapeGridData(GetGridData(81:82, 91:94, 1991:1996, c("TMIN", "TMAX","TAVG", "WIND", "VPRESS", "PREC", "RAD")))
print(identical(b1, b1b))
```

```{r}
print(length(b1))
print(names(b1))
print(sapply(b1, length))
print(sapply(b1, names, simplify = FALSE))
print(dim(b1[[1]]$"1991")) # same as: print(dim(b1[[1]][[1]]))
print(dimnames(b1[[1]]$"1991"))
```

```{r}
Rows <- as.integer(names(b1)) %/% 1000
Columns <- as.integer(names(b1)) %% 1000
print(cbind(Rows, Columns))
```


Getting data in format ready to be used with crop models by country: GetGridCropSim(county, period)
```{r}
# GetCountryCropSim
# 
b2 <- GetCountryCropSim("LU", c(1991, 1996))
```

```{r}
print(length(b2))
print(names(b2))
print(sapply(b2, length))
print(sapply(b2, names, simplify = FALSE))
print(dim(b2[[1]]$"1991")) # same as: print(dim(b2[[1]][[1]]))
print(dimnames(b2[[1]]$"1991"))
```

```{r}
Rows <- as.integer(names(b2)) %/% 1000
Columns <- as.integer(names(b2)) %% 1000
print(cbind(Rows, Columns))
```

