---
title: "UsingSpectator"
output:
  rmarkdown::html_vignette:
    df_print: paged
    toc: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Your Vignette Title}
  %\VignetteEncoding{UTF-8}    
params:
  extra1: 'style="background-color: #ECF0F1; padding: 1px; display: inline-block;"'
---

```{r label = "knitr options", include = FALSE}
knitr::opts_chunk$set(
    fig.width = 7, 
    fig.height = 4,
    out.width = "100%",
    fig.align = "center",
    collapse = TRUE,
    comment = "#>"
)
knitr::opts_knit$set(global.device = TRUE)
options(width = 999)
```

```{r label = "httptest_start", include = FALSE}
library(httptest)
httptest::start_vignette("UsingSpectator")
```



# Introduction

The spectator package for R was developed to allow access to 
[Spectator Earth](https://spectator.earth/) API from R. Spectator Earth offers a Web app providing Earth Observation imagery, mainly from open data satellites like the Sentinel and the Landsat family. These features are also exposed through an API, and the goal of the spectator package is to provide easy access to this functionality from R.

The main functions allow to retrieve the acquisition plans for Sentinel-1, Sentinel-2 and Landsat-8 satellites and to get the past or (near)future overpasses over an area of interest for these satellites. It is also possible to search the archive for available images over the area of interest for a given (past) period, get the URL links to download the whole image tiles, or alternatively to download the image for just the area of interest based on selected spectral bands.

One can also get a current position and trajectory for a much larger set of satellites.

Other functions might be added in subsequent releases of the package.

# API key
Some of the functions (mainly those specific to Sentinel and Landsat satellites) require to pass an API key as a parameter to the function (because the underlying API endpoint requires it). The API key is automatically generated for every registered user at https://app.spectator.earth. You can find it under 'Your profile' (bottom left button) and copy it to clipboard. The functions in the spectator package by default retrieve the API key from the environment variable "spectator_earth_api_key". You can choose any other way of providing it, but keep in mind that for security reasons it is NOT recommended to hard-code (include it as clear text) it in your scripts.

```{r label = "setup", include = FALSE}
library(spectator)
```

# Satellites

We can get the list of all the satellites referenced in the [Spectator Earth](https://spectator.earth/) database and their current positions.


```{r label = "get satellites"}
pos <- GetAllSatellites(positions = TRUE)
head(pos, n = 20L)
```
We can now plot the current positions of the satellites on the world map.

```{r label = "plot satellites", out.extra = params$extra1}
library(sf)
library(maps)
maps::map("world", fill = TRUE, col = "lightgrey", mar = rep(0.5, 4))
```

We shall show the open data satellites in green, and all the others in red.

```{r label = "satellite positions", out.extra = params$extra1}
plot(sf::st_geometry(subset(pos, open == TRUE)), add = TRUE, col = "green", pch = 15)
plot(sf::st_geometry(subset(pos, open == FALSE)), add = TRUE, col = "red", pch = 16)
```

We can also add satellite names as labels. In order to be able to read them, we will shift labels up a little bit.

```{r label = "satellite names", out.extra = params$extra1}
xy <- sf::st_coordinates(pos)
xy[, 2] <- xy[, 2] + 3 
text(xy, labels = pos$name, cex = 0.5)

```
```{r label = "turn off global device", include = FALSE}
knitr::opts_knit$set(global.device = FALSE)
```

# Trajectories

get trajectory and current position for a selected satellite


```{r label = "get trajectories"}
sat <- "SPOT-6"
traj <- GetTrajectory(satellite = sat)
pos1 <- GetSatellite(satellite = sat, positions = TRUE)
```


 do some nice graphs
```{r label = "plot trajectories", out.extra = params$extra1}
maps::map("world", fill = TRUE, col = "lightgrey", mar = c(0.5, 0.5, 3, 0.5))
plot(sf::st_geometry(traj), lwd = 2, col = "red", add = TRUE)
plot(sf::st_geometry(pos1), pch = 15, col = "green", cex = 1.5, add = TRUE)
title(main = sprintf("current %s trajectory & position", sat))
```

# Acquisition plans

examples for acquisition plans
works only for Sentinel-1, Sentinel-2 and Landsat-8 satellites

get plans for all eligible satellites for today

```{r label = "acquisition plans"}
plans <- GetAcquisitionPlan()
head(plans, n = 20L)
```

explore the content of the data frame, you'll see that the available attributes vary with the satellite

focus on Sentinel 2
day after tomorrow
```{r label = "acquisition plans Sentinel 2"}
sat <- c("Sentinel-2A", "Sentinel-2B")
day <- Sys.Date()  + 2
plan <- GetAcquisitionPlan(satellites = sat, date = day)
head(plan)
```


do some nice graphs
```{r label = "plot acquisition plans", out.extra = params$extra1}
library(maps)
maps::map("world", fill = TRUE, col = "lightgrey", mar = c(0.5, 0.5, 3, 0.5))
plot(sf::st_geometry(plan), border = "red", add = TRUE)
title(main = sprintf("%s acquisition plan for %s", paste(sat, collapse = "/"), day))
```

# Satellite overpasses

get the Luxembourg country shape as area of interest
```{r label = "Luxembourg boundary"}
roi <- system.file("extdata", "luxembourg.geojson", package = "spectator")
boundary <- sf::read_sf(roi, as_tibble = FALSE)
```

first check that we have the api_key
need api key; so get it first, all set it in code
```{r label = "API key"}
my_key <- Sys.getenv("spectator_earth_api_key")
```

look for Sentinel-2 A and B, use shorthand notation, default time frame
```{r label = "get Sentinel 2 overpasses"}
pass <- GetOverpasses(boundary, satellites = "S-2", acquisitions = TRUE, api_key = my_key)
```

do some nice graphs
```{r label = "plot Sentinel 2 overpasses", out.extra = params$extra1}
library(maps)
days <- range(as.Date(pass$date))
satellites <- sort(unique(pass$satellite))
maps::map(database = "world", region = c("Belgium", "Netherlands", "Germany", "Luxembourg",
            "France", "Switzerland"), col = "lightgrey", fill = TRUE, mar = c(0.5, 0.5, 4, 0.5))
plot(sf::st_geometry(boundary), add = TRUE, col = "red", border = FALSE)
plot(sf::st_geometry(pass), add = TRUE)
title(main = sprintf("%s overpasses for period %s", paste(satellites, collapse = "/"), 
                     paste(days, collapse = ":")))

```

look for Sentinel-1 A and B, use shorthand notation, last week  
```{r label = "get Sentinel 1 overpasses"}
pass <- GetOverpasses(boundary, satellites = "S-1",
                      days_before = 7, days_after = 0, acquisitions = TRUE)
head(pass)
```


do some nice graphs
```{r label = "plot Sentinel 1 overpasses", out.extra = params$extra1}
days <- range(as.Date(pass$date))
satellites <- sort(unique(pass$satellite))
maps::map(database = "world", region = c("Belgium", "Netherlands", "Germany", "Luxembourg",
            "France", "Switzerland"), col = "lightgrey", fill = TRUE, mar = c(0.5, 0.5, 4, 0.5))
plot(sf::st_geometry(boundary), add = TRUE, col = "red", border = FALSE)
plot(sf::st_geometry(pass), add = TRUE)
title(main = sprintf("%s overpasses for period %s", paste(satellites, collapse = "/"), 
                     paste(days, collapse = ":")))
```

# Plot overpasses calendar

Let's try to get as much information as we can about Sentinel-2A and Sentinel-2B overpasses of Luxembourg.

```{r label = "get Sentinel 2 overpasses max time frame"}
pass <- GetOverpasses(boundary, satellites = "S-2", acquisitions = FALSE, api_key = my_key,
                      days_before = 99, days_after = 99)
```

As the acquisitions take place only in the middle of the day, so we shall keep only the overpasses between 06:00 and 18:00.

```{r}
# we need these packages
library(sf)
library(lutz)
library(lubridate)

tz <- lutz::tz_lookup(sf::st_centroid(sf::st_geometry(boundary)), method = "accurate")
localtime <- lubridate::local_time(pass$date, tz = tz, units = "hours")
pass$acquisition <- localtime > 6 & localtime < 18
pass <- subset(pass, acquisition == TRUE)
```

We can now plot the nice calendar using the calendR package.

```{r}
# we need this package
library(calendR)

time_span <- range(as.Date(pass$date))
satellites <- sort(unique(pass$satellite))
first_date <- lubridate::floor_date(lubridate::ymd(time_span[1]), unit = "month") 
last_date <- lubridate::ceiling_date(lubridate::ymd(time_span[2]), unit = "month") - 1
days <- first_date:last_date
passes <- as.Date(pass$date)
events <- rep(NA, length(days))
events[passes - first_date + 1] <- pass$satellite

# will use English month and day of week labels 
ans <- Sys.setlocale(category = "LC_ALL", locale = "English")

calendR::calendR(start_date = first_date, end_date = last_date, 
                 title = "Sentinel 2 overpasses calendar",
                 special.days = events,
                 special.col = 2:(length(satellites) + 1),
                 col = "grey",
                 mbg.col = 4,               # Color of the background of the names of the months
                 months.col = "white",      # Color text of the names of the months        
                 bg.col = "#f4f4f4",        # Background color
                 start = "M",               # Start the weeks on Monday
                 legend.pos = "right")
```

# Show overpasses in a calendar app

We can also export the overpasses calendar as a calendar file (iCal or ICS file), which can be imported in any calendar management application. You could thus take the overpasses calender with you in the calendar app of you smartphone, for example. We shall use the calendar package to generate the iCal file.


```{r}
# we need this package
library(calendar)

N <- nrow(pass)
lst <- vector(mode = "list", length = N)
for (i in 1:nrow(pass)) {
    x <- pass[i, ]
    lst[[i]] <- calendar::ic_event(uid = calendar::ic_guid(), start_time = x$date, end_time = x$date, summary = x$satellite)
} 

cal <- calendar::ical(do.call(rbind, lst))
# cal[, "TZID"] <- tz
koords <- sf::st_coordinates(sf::st_centroid(sf::st_geometry(boundary, of_largest_polygon = FALSE)))
cal[, "LOCATION"] <- sprintf("(LAT:%f;LONG:%f)", koords[, "Y"], koords[, "X"])
calFile <- "overpasses.ics"
calendar::ic_write(cal, file = calFile)
```

```{r label = "Google calendar screenshot", include = FALSE}
fichier <- normalizePath("data/Calendar.jpg")
```

We can now import this file in our preferred calendar management application, here Google calendar.
![Google calendar](`r fichier`){width=100%}
A screen shot of Google calendar showing the Sentinel-2A and Sentinel-2B overpasses of Luxembourg.

# Searching for images

We can query the archived images catalog to find available images for an area of interest (AOI).
It can be done for images captured by Sentinel-1, Sentinel-2 and Landsat-8 satellites.
This functionality requires a valid API key. So we shall first get the API key from an environment variable or set it in any other suitable way.

```{r label = "API key 2"}
my_key <- Sys.getenv("spectator_earth_api_key")
```

Let's say that our area of interest (AOI) is the New York City Central Park. We shall first read the contour from a GeoJSON file.

```{r label = "Central Park boundary"}
roi <- system.file("extdata", "centralpark.geojson", package = "spectator")
boundary <- sf::read_sf(roi, as_tibble = FALSE)
```

Now we shall look for all Sentinel 2 images (both Sentinel-2A and Sentinel-2B) captured in May 2021.

```{r label = "serch Central Park images"}
catalog <- SearchImages(aoi = boundary, satellites = "S2", 
                        date_from = "2021-05-01", date_to = "2021-05-30", 
                        footprint = FALSE, api_key = my_key)
head(catalog, n = 20L)
```

# Imagery files
To access the imagery files we have to use their ids. We shall concentrate on the image with minimal cloud coverage and get the corresponding id.

```{r label = "minimal cloud coverage"}
best_id <- catalog[order(catalog$cloud_cover_percentage), ][1, "id"]
```

We can now retrieve the list of all available (downloadable) files for this id (the image with minimal cloud coverage).

```{r label = "list imagery files"}
images <- GetImageryFilesList(best_id, api_key = my_key)
head(images, n = 20L)
```
```{r label = "file size", include = FALSE}
size <- as.integer(subset(images, path == "B04.jp2", "size"))
```

We could now download the whole band 4 image tile for the image with the minimal cloud coverage using the code below.
*Warning*: this is a very **BIG** file (`r prettyNum(size, big.mark = " ")` bytes), so think twice before you start the download!

```{r label = "download the big file", eval = FALSE}
from <- paste0(catalog[order(catalog$cloud_cover_percentage), ][1, "download_url"], "B04.jp2")
to <- "B04.jp2"
library(httr)
resp <- httr::GET(url = from, query = list(api_key = my_key))
writeBin(httr::content(resp), con = to)
```

# High resolution images

We can also get a get the high resolution images of the AOI (here Central Park) using the image id from the catalog. We shall use once again the id that corresponds to the with the minimal cloud coverage. One can request one or three bands, to get the monochrome or color (RGB) image. The Sentinel 2 RGB bands are 4, 3, and 2, respectively (source: https://en.wikipedia.org/wiki/Sentinel-2#Spectral_bands).


```{r label = "high resolution image", eval = FALSE}
img <- GetHighResolutionImage(aoi = boundary, id = best_id, bands = c(4, 3, 2), 
                              width = 1024, height = 1024,
                              file = tempfile(pattern = "img", fileext = ".jpg"), 
                              api_key = my_key)
```
```{r label = "high resolution image file name", include = FALSE}
fichier <- normalizePath("data/CentralPark.jpg")
```

High resolution image
![Central Park image](`r fichier`){width=100%}
Sentinel-2 image of Central Park, New York City on May 12, 2021.
<br>
<br>
<br>
<br>
_Compiled on `r Sys.time()`._


```{r label = "httptest_end", include = FALSE}
httptest::end_vignette()
```
